---
title: "NLP Analysis of My Scientific Jounral Library"
author: "Joshua Cook"
date: "May 4, 2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")

library(jhcutils)
library(pubmed.mineR)
library(bibtex)
library(RefManageR)
library(tidygraph)
library(ggraph)
library(magrittr)
library(tidyverse)
```

In this analysis, I use 'Pubtator' to analysis my own biomedical article library.

## Data

I began by exporting the data for all of my journal articles in Mendeley as a '.bib' file format. This is available in "data/mendeley_library.bib". The ['bibtex'](https://cran.r-project.org/web/packages/bibtex/index.html) package was used to read in the file as a `bibentry` object.

```{r read_library, warning=FALSE, message=FALSE, cache=TRUE}
bib <- bibtex::read.bib("data/mendeley_library.bib")
class(bib)
```

After 'bibtex' automatically ignores incomplete references, there were `r length(bib)` entries. I took a look at the first few entries to get an idea of what I was dealing with.

```{r show_top, echo=FALSE}
bib[1:2]
```

The data was turned into a tibble for easier use.

```{r bib_to_df}
bib <- RefManageR::as.BibEntry(bib) %>%
    as.data.frame() %>% 
    as_tibble()
```

The data was now in a more managable format for data analysis.

```{r table_bib, echo = FALSE}
bib
```


## Accessing Pubtator

I used the ['pubmed.mineR](https://cran.r-project.org/web/packages/pubmed.mineR/index.html) package (here is a link to the original publication [Rani *et al.*, (2015)](https://www.ncbi.nlm.nih.gov/pubmed/26564970)) to pull information on the pubications from Pubtator.

The information can be gathered for a publication using `pubtator_function(pmid)`.

```{r get_pubtato}
# turns the pubtator results into a tidy tibble, indexed by PMID
pubtator_to_tibble <- function(pubtato) {
    pubtato <- unlist(pubtato)
    if (!"PMID" %in% names(pubtato)) return(NULL)
    pmid <- pubtato[["PMID"]]  # extract the PMID
    # create a tibble with a row for each field and value, indexed by PMID
    tib <- tibble(PMID = pmid,
                  field = names(pubtato),
                  val = pubtato) %>% 
        filter(field != "PMID") %>%  # remove PMID
        mutate(field = str_remove_all(field, "[:digit:]+"))
    return(tib)
}


pubtatoes <- purrr::map(bib$pmid, pubmed.mineR::pubtator_function) %>% 
    purrr::map(pubtator_to_tibble) %>% 
    bind_rows()
```

```{r show_pubtato, echo = FALSE}
pubtatoes
```

There were a few cleaning steps needed. These were completed in the following pipeline - I tried to add comments wherever I could.

```{r clean_pubtatoes}
replace_synonyms <- function(x) purrr::map_chr(x, replace_synonym)
replace_synonym <- function(x) {
    if (x %in% c("epidermal growth factor receptor",
                 "Epidermal growth factor receptor")) {
        return("EGFR")
    } else if (x == "beta-catenin") {
        return("CTNNB1")
    } else if (x %in% c("K-ras", "K-Ras", "Kras", "KRas")) {
        return("KRAS")
    } else if (x %in% c("N-ras", "N-Ras", "Nras", "NRas")) {
        return("NRAS")
    } else if (x %in% c("H-ras", "H-Ras", "Hras", "HRas")) {
        return("HRAS")
    } else if (x %in% c("p53")) {
        return("TP53")
    } else if (x %in% c("Akt")) {
        return("AKT")
    } else if (x %in% c("Raf")) {
        return("RAF")
    } else if (x == "E-cadherin") {
        return("CDH1")
    }
    return(x)
}

pubtatoes %<>% 
    mutate(val = str_remove_all(val, "\\(|\\)")) %>%  # remove parantheses
    mutate(val = replace_synonyms(val)) %>%
    unique()
```


## Simple Analysis

### Which genes do I read about most frequently?

Since I study the genetics of *KRAS*, I expect there to be a lot of papers about its interactors and other oncogenes in *RAS*-driven cancers (such as *APC* in CRC and *STK11* in NSCLC).

```{r clean_pubtato}
pubtatoes %>% 
    filter(field == "Genes") %>%
    mutate(val = forcats::fct_lump_min(val, min = 3)) %>%
    filter(val != "Other") %>%
    mutate(val = forcats::fct_infreq(val)) %>%
    ggplot() +
    geom_bar(aes(x = val)) +
    theme_classic() +
    scale_y_continuous(expand = expand_scale(mult = c(0, 0.05))) +
    theme(axis.text.x = element_text(angle = 65, hjust = 1),
          axis.title.x = element_blank()) +
    labs(y = "count",
         title = "Frequency of genes in articles")
```

As expected, there are a lot of genes in the MAPK pathway, PI3Kalpha-Akt, oncogenes, and tumor supressors. Funny enough, a small project that I am working on concerning Polycomb repressive complex 2 (PRC2) has made quite the impact as several of the subunits are in the list, including *JARID2*, *EZH2*, and *SUZ12*. Interestingly, though, PRC1 is on the list, but the term PRC2, is not. I wonder if this is just due to the differences in how people discuss the two related complexes in papers.

`#> TODO: more simple plots to explore the data further`


## Disease Network

For the files with at least one disease and other entity (gene, chemical, etc.), I drew a link between them to create a disease network for my library. I used the ['tidygraph'](https://cran.r-project.org/web/packages/tidygraph/index.html) package for handling the graph.

```{r}
# remove the NULL values of a list
remove_nulls <- function(x) {
    idx <- purrr::map_lgl(x, is.null)
    return(x[!idx])
}

# make a graph for an individual file
file_disease_gene_graph <- function(PMID, data) {
    # get disases and genes for the file
    diseases <- data %>% filter(field == "Diseases") %>% u_pull(val) %>% unlist()
    genes <- data %>% filter(field != "Diseases") %>% u_pull(val) %>% unlist()
    
    # check there are both dieases and genes
    if (length(diseases) < 1 | length(genes) < 1) {
        return(NULL)
    }
    
    # construct the graph
    gr <- expand.grid(diseases, genes) %>% 
        as_tibble() %>%
        mutate(pmid = !!PMID) %>%
        as_tbl_graph(directed = TRUE)
    return(gr)
}

pub_gr <- pubtatoes %>% 
    group_by(PMID) %>% 
    tidyr::nest() %>%
    pmap(file_disease_gene_graph) %>% 
    remove_nulls() %>% 
    recursive_graph_join()
```

```{r show_pubgr, echo=FALSE}
pub_gr
```

I then annotated the nodes as either a disease or gene.

```{r annotate_nodes, error=TRUE}
tt <- pubtatoes %>% select(field, val) %>% unique()
names(tt) <- c("type", "name")
pub_gr <- pub_gr %N>%
    left_join(tt, by = "name")
```

This resulted in an error. Normally, I would fix this and no one would see the error, but the cause was too good not to share...

I have used 'tidygraph' long enough to know that this error is thrown when merging with nodes when there are some nodes being mapped to multiple rows in the joining data table (`tt` in this case). Therefore, I isolated the values in `tt$name` with more than one entry.

```{r uncover_error}
tt %>%
    mutate(name = fct_lump_min(name, min=2)) %>%  # lump values with <2 entries
    filter(name != "Other") %>%  # filter out the renamed values
    arrange(name)  # arrange by `name` column
```

This means that Pubtator listed *APC* both as a disease and a gene. This is realy not a bad prediction considering that this gene oringially  recieved its name from the disease it was first discovered while studying: Adenomatous polyposis coli. However, the others are less forgivable. *Drosophila* is not a gene, it is a fruit fly,therefore, `"Species"` was the correct designation. IPMN stands for "intraductal papillary mucinous neoplasm" which is a benign tumor that can progress to pancreatic cancer (more information can be found at the [American Cancer Society's website](https://www.cancer.org/cancer/pancreatic-cancer/about/what-is-pancreatic-cancer.html)).

Finally, I manually removed the erroneous entries and made the join to complete the disease network.

```{r annotate_nodes2}
tt %<>%
    filter(!(type == "Diseases" & name == "APC")) %>% 
    filter(!(type == "Diseases" & name == "Drosophila"))%>%
    filter(!(type == "Chemicals" & name == "IPMN"))
pub_gr <- pub_gr %N>%
    left_join(tt, by = "name")
```

```{r show_pubgr2, echo=FALSE}
pub_gr
```

To begin visualizing the network, I plotted the largest component (which was almost all of the nodes). The nodes were colored by their type, and their size was associated with how many connections they made. The edge width and transparency was associated with the number of articles (PMIDs) that made the link.

```{r plot_pubgr, echo = FALSE}
type_pal <- c(
    "Diseases" = "tomato",
    "Genes" = "dodgerblue",
    "Chemicals" = "seagreen2",
    "Species" = "purple1"
)

type_size <- c(
    "Diseases" = 2,
    "Genes" = 1,
    "Chemicals" = 1,
    "Species" = 1
)

pub_gr %>% 
    jhcutils::get_giant_component() %>%
    convert(to_simple) %E>% 
    mutate(edge_weight = map_dbl(.tidygraph_edge_index, length)) %N>%
    mutate(node_deg = centrality_degree(mode = "all")) %>%
    ggraph(layout = "kk") +
    geom_edge_link(aes(width = edge_weight, alpha = edge_weight),
                       color = "grey50") +
    geom_node_point(aes(color = type, size = node_deg)) +
    scale_color_manual(values = type_pal) +
    scale_edge_width(range = c(1, 3)) +
    theme_void() +
    labs(color = "",
         size = "degree",
         edge_width = "no. PMIDs", edge_alpha = "no. PMIDs",
         title = "Disease network of my article library")
```

A few things stood out to me. First, the network is incredibly dense. This is likely caused by the largse number of disease being highly redundant (ie. "colorectal cancer", "COAD", "CRC"). These would likely share a lot of the other entities, pulling the network tighter.

Also, there are *way* more species than I would have predicted. **These warrant further investigation**.

Considering only the diseases and genes

```{r plot_pubgr2, echo = FALSE}
pub_gr %>% 
    jhcutils::get_giant_component() %>%
    filter(type %in% c("Diseases", "Genes")) %>%
    filter(centrality_degree(mode = "all") > 0) %>%
    convert(to_simple) %E>% 
    mutate(edge_weight = map_dbl(.tidygraph_edge_index, length)) %N>%
    mutate(node_deg = centrality_degree(mode = "all")) %>%
    ggraph(layout = "kk") +
    geom_edge_link(aes(width = edge_weight, alpha = edge_weight),
                       color = "grey50") +
    geom_node_point(aes(color = type, size = node_deg)) +
    scale_color_manual(values = type_pal) +
    scale_edge_width(range = c(1, 3)) +
    theme_void() +
    labs(color = "",
         size = "degree",
         edge_width = "no. PMIDs", edge_alpha = "no. PMIDs",
         title = "Disease network of my article library",
         subtitle = "showing only the diseases and genes")

```

