---
title: "NLP Analysis of My Scientific Jounral Library"
author: "Joshua Cook"
date: "May 4, 2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")

library(pubmed.mineR)
library(bibtex)
library(RefManageR)
library(magrittr)
library(tidyverse)
```

In this analysis, I use 'Pubtator' to analysis my own biomedical article library.

## Data

I began by exporting the data for all of my journal articles in Mendeley as a '.bib' file format. This is available in "data/mendeley_library.bib". The ['bibtex'](https://cran.r-project.org/web/packages/bibtex/index.html) package was used to read in the file as a `bibentry` object.

```{r read_library, warning=FALSE, message=FALSE, cache=TRUE}
bib <- bibtex::read.bib("data/mendeley_library.bib")
class(bib)
```

After 'bibtex' automatically ignores incomplete references, there were `r length(bib)` entries. I took a look at the first few entries to get an idea of what I was dealing with.

```{r show_top, echo=FALSE}
bib[1:2]
```

The data was turned into a tibble for easier use.

```{r bib_to_df}
bib <- RefManageR::as.BibEntry(bib) %>%
    as.data.frame() %>% 
    as_tibble()
```

The data was now in a more managable format for data analysis.

```{r table_bib, echo = FALSE}
bib
```


## Accessing Pubtator

I used the ['pubmed.mineR](https://cran.r-project.org/web/packages/pubmed.mineR/index.html) package (here is a link to the original publication [Rani *et al.*, (2015)](https://www.ncbi.nlm.nih.gov/pubmed/26564970)) to pull information on the pubications from Pubtator.

The information can be gathered for a publication using `pubtator_function(pmid)`.

```{r get_pubtato}
# turns the pubtator results into a tidy tibble, indexed by PMID
pubtator_to_tibble <- function(pubtato) {
    pubtato <- unlist(pubtato)
    if (!"PMID" %in% names(pubtato)) return(NULL)
    pmid <- pubtato[["PMID"]]  # extract the PMID
    # create a tibble with a row for each field and value, indexed by PMID
    tib <- tibble(PMID = pmid,
                  field = names(pubtato),
                  val = pubtato) %>% 
        filter(field != "PMID") %>%  # remove PMID
        mutate(field = str_remove_all(field, "[:digit:]+"))
    return(tib)
}


pubtatoes <- purrr::map(bib$pmid, pubmed.mineR::pubtator_function) %>% 
    purrr::map(pubtator_to_tibble) %>% 
    bind_rows()
```

```{r show_pubtato, echo = FALSE}
pubtatoes
```

There were a few cleaning steps needed. These were completed in the following pipeline - I tried to add comments wherever I could.

```{r clean_pubtatoes}
replace_synonyms <- function(x) purrr::map_chr(x, replace_synonym)
replace_synonym <- function(x) {
    if (x %in% c("epidermal growth factor receptor",
                 "Epidermal growth factor receptor")) {
        return("EGFR")
    } else if (x == "beta-catenin") {
        return("CTNNB1")
    } else if (x %in% c("K-ras", "K-Ras", "Kras", "KRas")) {
        return("KRAS")
    } else if (x %in% c("N-ras", "N-Ras", "Nras", "NRas")) {
        return("NRAS")
    } else if (x %in% c("H-ras", "H-Ras", "Hras", "HRas")) {
        return("HRAS")
    } else if (x %in% c("p53")) {
        return("TP53")
    } else if (x %in% c("Akt")) {
        return("AKT")
    } else if (x %in% c("Raf")) {
        return("RAF")
    } else if (x == "E-cadherin") {
        return("CDH1")
    }
    return(x)
}

pubtatoes %<>% 
    mutate(val = str_remove_all(val, "\\(|\\)")) %>%  # remove parantheses
    mutate(val = replace_synonyms(val)) %>%
    unique()
```


## Simple Analysis

### Which genes do I read about most frequently?

Since I study the genetics of *KRAS*, I expect there to be a lot of papers about its interactors and other oncogenes in *RAS*-driven cancers (such as *APC* in CRC and *STK11* in NSCLC).

```{r}
pubtatoes %>% 
    filter(field == "Genes") %>%
    mutate(val = forcats::fct_lump_min(val, min = 3)) %>%
    filter(val != "Other") %>%
    mutate(val = forcats::fct_infreq(val)) %>%
    ggplot() +
    geom_bar(aes(x = val)) +
    theme_classic() +
    scale_y_continuous(expand = expand_scale(mult = c(0, 0.05))) +
    theme(axis.text.x = element_text(angle = 65, hjust = 1),
          axis.title.x = element_blank()) +
    labs(y = "count",
         title = "Frequency of genes in articles")
```

As expected, there are a lot of genes in the MAPK pathway, PI3Kalpha-Akt, oncogenes, and tumor supressors. Funny enough, a small project that I am working on concerning Polycomb repressive complex 2 (PRC2) has made quite the impact as several of the subunits are in the list, including *JARID2*, *EZH2*, and *SUZ12*. Interestingly, though, PRC1 is on the list, but the term PRC2, is not. I wonder if this is just due to the differences in how people discuss the two related complexes in papers.

